~\Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores

### 2024-07-24 manuscript/paper.tex
Update Paper
Yoel adding text to introduction, updating and formatting references, etc.

\title{~\Large Increased sexual dimorphism evolves in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
Raheyma Siddiqui\(^1\), Akhil Ghosh\(^2\)\\
\large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
Matthews\(^{2,3}\), and Yoel E. Stuart\(^{1,+}\)\\
\vspace{-1.1mm}\\
\large \(^1\) Department of Biology, Loyola University Chicago, Chicago,
IL, USA \vspace{-1.1mm}\\
\large \(^2\) Department of Mathematics and Statistics, Loyola
University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^3\) Center for Data Science and Consulting, Loyola University
Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^4\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\
\large \(^*\) Equal contribution \vspace{-1.1mm}\\
\large \(^+\) Corresponding:
\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & stl & Distance from anterior tip of premaxilla to
posterior end of last vertebra (hypural plate) \\
Dorsal Spine & mds & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & mdf & Number of bones in the dorsal fin posterior to
the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & maf & Number of bones in the anal fin posterior to the
anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & mcv & Number of vertebrae posterior to and including
the first vertebra contacting an anal fin pterygiophore (Aguirre et
al.~2014) \\
Pterygiophore number & mpt & Number of pterygiophores anterior to but
excluding the pterygiophore under the third dorsal spine, which is
immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & lps.sc & Length from the base of one pelvic spine
above its articulation with the pelvic girdle to its distal tip \\
Ectocoracoid & ect.sc & Length between the anterior and posterior tips
of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & tpg.sc & Length between the anterior to posterior tips
along midline. If vestigial, the sum of longest anterior-posterior axis
for the vestiges \\
Cleithrum length & cle.sc & Length from free dorsal tip to ventral tip
of the cleithrum on the anterior margin of the shoulder girdle (i.e.,
cleithrum) \\
Premaxilla & pmx.sc & Length from the anterior tip of the premaxilla to
the distal tip of the ascending process of the premaxilla \\
Dorsal Spine & Ds\#.sc\# = 1,2,or 3 & Length from the base of a dorsal
spine above the pterygiophore to its distal tip along the anterior
edge \\
Pterygiophore & lpt.sc & Distance between the anterior to posterior tips
of the pterygiophore immediately preceding the 3rd dorsal spine (when
present) \\
\bottomrule()
\end{longtable}

@@

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

In addition, we assume that the discrete traits abdominal vertebrae
(mav) and caudal vertebrae (mcv) also have correlation with the standard
length of the fish (\textbf{CITATION}). If \(Y_{k,ij}\) is one of the
above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we also assume the above model except we
set \(\gamma_k = 0\) because of the assumption of no correlation between
the standard length and these traits. More specifically, we assume
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} Poisson(\exp\{\mu_{k,ft_i}\}), & W_{ij} = \text{Female} \\ Poisson(\exp\{\mu_{k,mt_i}\}), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}

In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
model the time-\(t_i\) specific mean of phenotype \(k\) for female and
male stickleback fish, respectively. We point out that, for the discrete
traits, the means are represented by \(\exp\{\mu_{k,ft_i}\}\) and
\(\exp\{\mu_{k,mt_i}\}\) for ease of use in our modeling technique. We
further set \begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\) where \(\beta_{0,kg}\) and
\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
gender, accounting for the possibility of a time-dependent trend in the
mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
account for potential correlations between the residuals for a given
trait \(k\) and gender \(g\), we fit an Ornstein-Uhlenback (OU) process
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
change in \(u_k,gt\) for a given trait \(k\) and gender \(g\) over a
miniscule time period \(dt\). The OU process is defined as \begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align} where \(\kappa_k\) is a parameter associated with the
correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
the standard deviation of the OU process, and \(W_t\) is a standard
Brownian motion. As shown in (Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})), the closed form solution for
the SDE in (\ref{eq:OU_cont}) is \begin{align}
u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
constant, and we discuss the procedure for estimating this value below.

@@

Because we are fitting a dataset with a stochastic structure on the
means of the phenotypes, we analyze the data via a Bayesian analysis.
Bayesian data analysis is also more naturally used when we have to
impute data (\textbf{CITATION}).

Priors: For \(k = 1,\cdots,K\),

\begin{align}
u_{k,gt_1} & \overset{iid}{\sim}\mathcal{N}(0,\tau_{0,k}) \nonumber \\
\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
\tau_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
\tau_{0,k} & \overset{iid}{\sim}\mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
\gamma_{k} & \overset{iid}{\sim}\mathcal{N}(0,5) \nonumber \\
\beta_{0,kg} & \overset{iid}{\sim}\mathcal{N}(0,100) \nonumber \\
\beta_{1,kg} & \overset{iid}{\sim}\mathcal{N}(0,3) \nonumber \\
\label{eq:priors}
\end{align} We also note that, for the discrete phenotypes in equations
(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no \(\sigma_k\),
and thus is not sampled for those phenotypes.

@@

Bayesian Analysis after multiple imputation Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2009}{2010}): They recommend using a
large number of imputations. 5 or 10 is too small. We are using M = 100.

@@

>>>>

\title{~\Large Increased sexual dimorphism evolves in a fossil
stickleback following ecological release from fish piscivores}
\author{\large Allison Ozark\(^{1,*}\), Matthew Stuart\(^{2,3,*}\),
Raheyma Siddiqui\(^1\), Akhil Ghosh\(^{2,3}\)\\
\large Samantha Swank\(^1\), Michael A. Bell\(^4\), Gregory J.
Matthews\(^{2,3}\), and Yoel E. Stuart\(^{1,+}\)\\
\vspace{-1.1mm}\\
\large \(^1\) Department of Biology, Loyola University Chicago, Chicago,
IL, USA \vspace{-1.1mm}\\
\large \(^2\) Department of Mathematics and Statistics, Loyola
University Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^3\) Center for Data Science and Consulting, Loyola University
Chicago, Chicago, IL, USA \vspace{-1.1mm}\\
\large \(^4\) University of California Museum of Paleontology, Berkeley,
CA, USA \vspace{-1.1mm}\\
\large \(^*\) Equal contribution \vspace{-1.1mm}\\
\large \(^+\) Corresponding:
\href{mailto:ystuart@luc.edu}{\nolinkurl{ystuart@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3492}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1905}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4603}}@{}}
\caption{Traits and trait descriptions. `sc' denotes size correction of
trait against standard length. Names of bones follow Bowne
(\protect\hyperlink{ref-Bowne1994}{1994}) unless otherwise
noted.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Trait Name
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Code
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Trait Description
\end{minipage} \\
\midrule()
\endhead
Standard Length & \texttt{stl} & Distance from anterior tip of
premaxilla to posterior end of last vertebra (hypural plate) \\
Dorsal Spine & \texttt{mds} & Number of dorsal spines from 0 to 3 \\
Dorsal Fin Ray & \texttt{mdf} & Number of bones in the dorsal fin
posterior to the third dorsal spine (i.e., soft dorsal fin rays) \\
Anal Fin Ray & \texttt{maf} & Number of bones in the anal fin posterior
to the anal spine (i.e., soft dorsal fin rays) \\
Abdominal Vertebra & mav & Number of vertebrae anterior to the first
vertebra contacting an anal fin pterygiophore (Aguirre et al.~2014) \\
Caudal Vertebra & \texttt{mcv} & Number of vertebrae posterior to and
including the first vertebra contacting an anal fin pterygiophore
(Aguirre et al.~2014) \\
Pterygiophore number & \texttt{mpt} & Number of pterygiophores anterior
to but excluding the pterygiophore under the third dorsal spine, which
is immediately anterior to and contiguous with the dorsal fin \\
Pelvic Spine length & \texttt{lps.sc} & Length from the base of one
pelvic spine above its articulation with the pelvic girdle to its distal
tip \\
Ectocoracoid & \texttt{ect.sc} & Length between the anterior and
posterior tips of the shoulder girdle base (i.e., ectocoracoid) \\
Pelvic Girdle & \texttt{tpg.sc} & Length between the anterior to
posterior tips along midline. If vestigial, the sum of longest
anterior-posterior axis for the vestiges \\
Cleithrum length & \texttt{cle.sc} & Length from free dorsal tip to
ventral tip of the cleithrum on the anterior margin of the shoulder
girdle (i.e., cleithrum) \\
Premaxilla & \texttt{pmx.sc} & Length from the anterior tip of the
premaxilla to the distal tip of the ascending process of the
premaxilla \\
Dorsal Spine & \texttt{Ds\#.sc\#\ =\ 1,2,or\ 3} & Length from the base
of a dorsal spine above the pterygiophore to its distal tip along the
anterior edge \\
Pterygiophore & \texttt{lpt.sc} & Distance between the anterior to
posterior tips of the pterygiophore immediately preceding the 3rd dorsal
spine (when present) \\
\bottomrule()
\end{longtable}

@@

We impute the missing gender for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

@@

If \(Y_{k,ij}\) is a discrete trait, the conventional method of
modelling this data is by fitting a Poisson distribution. However, the
Poisson distribution assumes that the mean and variance of \(Y_{k,ij}\)
are equal (\(E(Y_{k,ij}) = \text{Var}(Y_{k,ij})\)), while the empirical
fossil data may have variances that are smaller than their respective
means. To combat this issue, we propose to fit the discrete traits to a
generalized Poisson model as defined in
(\protect\hyperlink{ref-GeneralizedPoisson}{\textbf{GeneralizedPoisson?}}).
Specifically, if \(X \sim GP(\lambda,\alpha)\), then \begin{align}
P(X = x) = \left\{\begin{array}{cc} \frac{(1 - \alpha)\lambda[(1 - \alpha)\lambda + \alpha x]^{x - 1} \exp\left\{-((1 - \alpha)\lambda  + \alpha x)\right\}}{x!} & (1 - \alpha)\lambda  + \alpha x \geq 0  \\ 0 & (1 - \alpha)\lambda  + \alpha x < 0 \end{array}\right.,
\label{eq:GP_pmf}
\end{align} where \(E(X) = \lambda\) and
\(Var(X) = \frac{\lambda}{(1 - \alpha)^2}\). If \(\alpha > 0\), then the
variance is greater than the mean, called overdispersion; if
\(\alpha < 0\), then the variance is greater than the mean, called
underdispersion; if \(\alpha = 0\), then the model degenerates to a
Poisson distribution.

We will use the generalized Poisson to model all of the discrete traits
in our dataset. In addition, we assume that the discrete traits
abdominal vertebrae (mav) and caudal vertebrae (mcv) also have
correlation with the standard length of the fish (\textbf{CITATION}). If
\(Y_{k,ij}\) is one of the above traits, then \begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i} + \gamma_kY_{K,ij}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_corr}
\end{align}

For the other discrete traits, we also assume the above model except we
set \(\gamma_k = 0\) because of the assumption of no correlation between
the standard length and these traits. More specifically, we assume
\begin{align}
{Y}_{k,ij} & \sim \left\{\begin{array}{ll} GP(\exp\{\mu_{k,ft_i}\},\alpha_k), & W_{ij} = \text{Female} \\ GP(\exp\{\mu_{k,mt_i}\},\alpha_k), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:disc_ind}
\end{align}

In the above model descriptions, \(\mu_{k,ft_i}\) and \(\mu_{k,mt_i}\)
model the time-\(t_i\) specific mean of phenotype \(k\) for female and
male stickleback fish, respectively. We point out that, for the discrete
traits, the means are represented by \(\exp\{\mu_{k,ft_i}\}\) and
\(\exp\{\mu_{k,mt_i}\}\) for ease of use in our modeling technique. We
further set \begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align} for \(g \in \{f,m\}\) where \(\beta_{0,kg}\) and
\(\beta_{1,kg}\) are regression parameters of phenotype \(Y_k\) for each
gender, accounting for the possibility of a time-dependent trend in the
mean structure, and \(u_{k,gt_i}\) is the corresponding residual. To
account for potential correlations between the residuals for a given
trait \(k\) and gender \(g\), we fit an Ornstein-Uhlenback (OU) process
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})). More
specifically, define \(du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}\), the
change in \(u_k,gt\) for a given trait \(k\) and gender \(g\) over a
miniscule time period \(dt\). The OU process is defined as \begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align} where \(\kappa_k\) is a parameter associated with the
correlation between \(u_{k,gt}\) and \(u_{k,g(t+dt)}\), \(\tau_k\) is
the standard deviation of the OU process, and \(W_t\) is a standard
Brownian motion. As shown in Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930}), the closed form solution for
the SDE in (\ref{eq:OU_cont}) is \begin{align}
u_{k,gt_i} \overset{iid}{\sim}\mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align} for \(i = 2,\cdots,T\). In a traditional OU process, the
initial value \(u_{k,gt_1}\) is assumed to be a (potentially unknown)
constant, and we discuss the procedure for estimating this value below.

@@

Because we are fitting a dataset with a stochastic structure on the
means of the phenotypes, and we are interested in the distribution of
the overall structure of these means, we analyze the data via a Bayesian
analysis. Bayesian data analysis is also more naturally used when we
have to impute data (\textbf{CITATION}). To aid in the sampling
procedure for the discrete phenotypes, we declare

Priors: To aid in the sampling procedure for the discrete phenotypes, we
declare
\(\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)\)
where
\(\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.\).
This transformation is performed to ensure our algorithm can properly
sample from the posterior distribution of interest.

For \(k = 1,\cdots,K\),

\begin{align}
u_{k,gt_1} & \overset{iid}{\sim}\mathcal{N}(0,\tau_{0,k}) \nonumber \\
\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\sigma > 0\}} \nonumber \\
\sigma_k & \overset{iid}{\sim}\mathcal{N}(0,10) \nonumber \\
\tau_k & \overset{iid}{\sim}\mathcal{N}(0,10)I_{\{\tau > 0\}} \nonumber \\
\tau_{0,k} & \overset{iid}{\sim}\mathcal{N}(0,20)I_{\{\tau > 0\}} \nonumber \\
\kappa_k & \overset{iid}{\sim}\mathcal{N}(0,1)I_{\{\kappa_g > 0\}} \nonumber \\
\gamma_{k} & \overset{iid}{\sim}\mathcal{N}(0,5) \nonumber \\
\beta_{0,kg} & \overset{iid}{\sim}\mathcal{N}(0,100) \nonumber \\
\beta_{1,kg} & \overset{iid}{\sim}\mathcal{N}(0,3) \nonumber \\
\label{eq:priors}
\end{align} We also note that, for the discrete phenotypes in equations
(\ref{eq:disc_corr}) and (\ref{eq:disc_ind}), there is no \(\sigma_k\),
and is not sampled, and for the continuous phenotypes in equation
(\ref{eq:cont}), there is no \(\phi_k\) and is not sampled.

@@

@@

As previously described, four models for the mean process were
considered for each trait. Within each trait, the model with the lowest
DIC was chosen for that train. All subsequent results presented here use
the model with the lowest DIC individually by trait. Table
\ref{whichmodel} shows the model that was selected for each trait. About
half of the traits had models with a trend, and most models did not
include an OU component. The traits that did include an OU component
were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal
vertebra (mav), and standard length (stl) with caudal vertebra (mcv)
being the only model with both a trend and an OU component. This implies
that not only does caudal vertebra size have a linear trend across
gender, but the average size at one time period is highly correlated
with the average size at the next time period.

\hypertarget{best-fitting-models}{%
\subsection{Best fitting models}\label{best-fitting-models}}

\begin{table}[ht]
\centering
\begin{tabular}{|c||c|c|}
  \hline
 & OU & No OU  \\ 
  \hline
  \hline
Trend  &  mcv  & ds2, ds3, ect, maf, mdf \\ 
  \hline
  No Trend & lpt, mav, stl  &  cle, ds1, lps, pmx\\
   \hline
\end{tabular}
\label{whichmodel}
\end{table}

\hypertarget{sexual-dimorphism}{%
\subsection{Sexual Dimorphism}\label{sexual-dimorphism}}

Table \ref{tab:prob0-table} shows the posterior probability of the
differences in the means between male and female specimens at each time
point. Probabilities near 0.5 indicate very little difference in the
means whereas probabilities far from 0.5 indicate sexual dimorphism with
values of 0 and 1 indicating larger mean values for females and males,
respectively. These results are shown graphically in figure
\ref{fig:prob0-plot}.

Figure \ref{fig:mu-diff-plot} shows the posterior mean difference for
males versus females (values above 0 indicate means that were greater
for males versus females). The full poosterior distributions over time
for all traits are shown in the appendix.

Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray
(mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant
sexual dimorphism with all but the last having learger mean values in
male sepcimen. Other traits such as abdominal vertebra (mav), caudal
vertebra (mcv), and pelvic spine length (lps) don't appear to have any
discernable sexual dimorphism in the means. Dorsal spine 3 (ds3)
exhibits sexual dimorphism that changes over time. The mean number of
dorsal spine 3 begins larger for females for the earliest observed time
periods. Over time, however, this switches and the male specimens
exhibit larger mean values for the latest observed time periods.

\hypertarget{changes-over-time}{%
\subsection{Changes over time}\label{changes-over-time}}

Table \ref{tab:auc-table} shows the probability that the posterior
distribution of the differences in the mean is greater than the
posterior distribution of the differences at the first time point.
Values near 0.5 indicate very little difference between the
distributions whereas values closer to 0 and 1 indicate changes in the
distributions with values of 1 meaning the difference in means has
shifted towards males having a larger mean and values of 0 indicating
the difference in means has shifted towards larger means in females.
These results are shown graphically in figure \ref{fig:AUC-plot}.

The amount of sexual dimporphism over time does not change substantially
for a majority of the trait observed here (i.e.~cle, ds1, ds2, ect, lps,
lpt, mav, mcv, pmx, stl). Traits that trend away from the mean
difference at the first observed time point include dorsal spine 3
(ds3), anal fin ray (maf), and dorsal fin ray (mdf).

The posterior mean of the mean number of dorsal spine rays at the first
observed time point is 2.29 for males and for females it is 2.45. At the
last observed time point, the average number for males and female both
dropped to 1.265 and 1.185 for males and females, respectively. This
resulted in a mean number of dorsal spine 3 lost by males of 1.027 and
for females it was 1.264. This resulted in a change of the mean
difference from time point 1 to time point 18 of 0.236. Basically, males
and females both lowered the mean number of dorsal spine 3, but females
lowered their mean more than the males.

We also observed a change in the differences between males and females
for the number of bones int he anal fin ray (maf) and the number of
bones in the dorsal fin ray. The difference in the means of males versus
females for anal fin ray bones changed by 0.152 over the course of the
observed time periods. This change was driven mostly by the change in
females. At the first observed time period the had on average 8.291
bones dropping to 8.170 over the 18 times periods observed. Whereas for
males the change in means rose only from 8.518 to 8.549, about 0.0316
bones on average.

A similar change in the difference of the means over time was observed
for the number of bones in the dorsal fin (mdf). However, unlike maf,
this change in sexual dimporhism was driven entirely by the males. The
mean change for females over the 18 observed time periods was only
0.0055 with the average number of dorsal fin bones for females starting
at 9.287 and the final observed mean of 9.281 resulting in almost no
change in the mean. For males, the average number of dorsal fin bones at
the first time point was about 9.568 dropping to 9.734 over the course
of the observed time periods for a gain of 0.166 bones on average.

\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedleft
time
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
cle
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ds1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ds2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ds3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ect
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
lps
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
lpt
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
maf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
mav
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
mcv
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
mdf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
pmx
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
stl
\end{minipage} \\
\midrule()
\endhead
1 & 0.913 & 0.215 & 0.359 & 0.203 & 0.865 & 0.428 & 0.853 & 0.944 &
0.391 & 0.507 & 0.959 & 0.948 & 0.883 \\
2 & 0.942 & 0.212 & 0.325 & 0.180 & 0.891 & 0.404 & 0.853 & 0.948 &
0.520 & 0.520 & 0.963 & 0.967 & 0.908 \\
3 & 0.843 & 0.207 & 0.249 & 0.153 & 0.770 & 0.279 & 0.803 & 0.965 &
0.289 & 0.525 & 0.977 & 0.912 & 0.777 \\
4 & 0.890 & 0.210 & 0.249 & 0.201 & 0.830 & 0.372 & 0.741 & 0.977 &
0.435 & 0.589 & 0.986 & 0.935 & 0.831 \\
5 & 0.954 & 0.211 & 0.211 & 0.195 & 0.909 & 0.377 & 0.756 & 0.985 &
0.533 & 0.622 & 0.992 & 0.975 & 0.921 \\
6 & 0.885 & 0.211 & 0.134 & 0.171 & 0.820 & 0.291 & 0.686 & 0.991 &
0.645 & 0.456 & 0.996 & 0.949 & 0.811 \\
7 & 0.812 & 0.211 & 0.124 & 0.239 & 0.760 & 0.340 & 0.617 & 0.994 &
0.224 & 0.717 & 0.999 & 0.879 & 0.747 \\
8 & 0.928 & 0.210 & 0.135 & 0.258 & 0.885 & 0.387 & 0.706 & 0.994 &
0.342 & 0.730 & 1.000 & 0.969 & 0.872 \\
9 & 0.899 & 0.213 & 0.145 & 0.304 & 0.857 & 0.406 & 0.715 & 0.995 &
0.521 & 0.756 & 1.000 & 0.940 & 0.845 \\
10 & 0.954 & 0.217 & 0.235 & 0.555 & 0.940 & 0.571 & 0.810 & 0.997 &
0.454 & 0.719 & 1.000 & 0.976 & 0.935 \\
11 & 0.913 & 0.211 & 0.162 & 0.476 & 0.875 & 0.388 & 0.703 & 0.997 &
0.363 & 0.657 & 1.000 & 0.957 & 0.864 \\
12 & 0.949 & 0.221 & 0.262 & 0.726 & 0.929 & 0.585 & 0.803 & 0.997 &
0.462 & 0.595 & 1.000 & 0.968 & 0.926 \\
13 & 0.904 & 0.220 & 0.260 & 0.789 & 0.885 & 0.598 & 0.795 & 0.998 &
0.470 & 0.370 & 1.000 & 0.929 & 0.872 \\
14 & 0.925 & 0.220 & 0.280 & 0.883 & 0.906 & 0.659 & 0.854 & 0.997 &
0.526 & 0.472 & 1.000 & 0.958 & 0.898 \\
15 & 0.880 & 0.225 & 0.283 & 0.839 & 0.849 & 0.617 & 0.796 & 0.997 &
0.490 & 0.398 & 1.000 & 0.919 & 0.834 \\
16 & 0.953 & 0.216 & 0.276 & 0.931 & 0.931 & 0.614 & 0.831 & 0.997 &
0.414 & 0.541 & 1.000 & 0.975 & 0.925 \\
17 & 0.803 & 0.219 & 0.232 & 0.832 & 0.765 & 0.383 & 0.655 & 0.996 &
0.569 & 0.758 & 1.000 & 0.854 & 0.751 \\
18 & 0.780 & 0.212 & 0.225 & 0.817 & 0.726 & 0.297 & 0.558 & 0.995 &
0.705 & 0.722 & 0.999 & 0.859 & 0.717 \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0602}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 26\tabcolsep) * \real{0.0723}}
